---
title: "Data Analysis - Daily Steps Takes"
author: "Jorge Espina"
date: "2/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , cache = FALSE)
```
1. Loading the data
-------------------

We set the working directory followed by downloading the dataset from its url and unzipping the file to "step_data.csv". The data comes from Roger Pengs [github account](https://github.com/rdpeng/RepData_PeerAssessment1).
```{r load data}
unzip(zipfile = "activity.zip")
unlink("activity.zip")

library(ggplot2)
library(plyr)
library(scales)
library(lattice) 
activity <- read.csv("activity.csv")

activity$date <- as.Date(activity$date, "%Y-%m-%d")
activity <- as.data.frame(activity)

```

2. Mean of "total number of step taken per day" over all days
-------------------------------------------------------------
Calculate the total number of steps taken per day and histogram Total Steps per Day

```{r  total number of step raken day}

steps <- with(activity, tapply(steps, date, sum, na.rm = TRUE))
hist(steps, breaks=6, xlab="Steps", main = "Total Steps per Day")


```
Media
```{r  media}
mean(steps) 


```
Mediana
```{r  mediana}

median(steps)

```
3. Average daily activity pattern
-------------------------------------------------------------
In this section, we make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken averaged across all days.
```{r  averange daily activity }
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")

clean <- activity[!is.na(activity$steps),]

```
create average number of steps per interval
```{r  create number of steps interval}
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))


```
Create line plot of average number of steps per interval
```{r  Average Number of Steps per Interval}
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")

```

Maximum steps by interval
```{r  maximum steps by interval}
maxSteps <- max(intervalTable$Avg)


```

Which interval contains the maximum average number of steps
```{r  interval contains the max}

intervalTable[intervalTable$Avg==maxSteps,1]

```
4. Imputing missing values
--------------------------

First of all, let us get a sense for the missing values. Are there days with all time intervals reporting NA step values? 
We can replace the missing data for a day by the time average over all other days.

```{r  dataac}
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
clean <- activity[!is.na(activity$steps),]
nrow(activity[is.na(activity$steps),])
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))
nadata<- activity[is.na(activity$steps),]
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")
mergeData <- rbind(clean, newdata2)
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")

## Mean of Steps with NA data taken care of
as.integer(mean(sumTable2$Steps))

## Median of Steps with NA data taken care of
as.integer(median(sumTable2$Steps))


```
Creating the histogram of total steps per day, categorized by data set to show impact

```{r  histogram of total steps per day, categorized by data set to show impact}
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Blue")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Orange", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("blue", "orange") )

```

5. Differences in activity patterns: Weekdays vs Weekends
---------------------------------------------------------

Create a new column describing if the date is a weekday or weekend.

Create new category based on the days of the week
```{r  create new category}

mergeData$DayCategory <- ifelse(mergeData$day %in% c("sabado", "domingo"),"Weekend","Weekday")
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

```

Creating Plot Average Steps per Interval Based on Type of Day

```{r  plot}

xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",layout = c(0,2),main="Average Steps per Interval Based on Type of Day"
       ,ylab="Average Number of Steps", xlab="Interval")
```